---
title: "Homework Assignment#2-Explore Gapminder and use dplyr"
output: 
  html_document:
    keep_md: true
        
---



## R Markdown
## Step1: Loading Gapminder and Tidyverse 
```{r}
library(gapminder)
library(tidyverse)
library(dplyr)
```



## Step2: Tests on the gapminder data frame
```{r}
typeof(gapminder)
class(gapminder)
ncol(gapminder)
nrow(gapminder)
dim(gapminder)
length(gapminder)
str(gapminder)
sapply(gapminder, class)
sapply(gapminder, typeof)

```

#### The above output gives the basic information about the data frame.
#### typof() function gives the data type of the function and in this case it is a "List"
#### nrow() and ncol() functions specify the number of rows and coloumn in the data frame. the output shows that the above data frame has "6 columns and 1704 rows"
#### dim() represents the dimension of the data frame in the form of matrix that is as rows and columns
#### str() generates the structure of the entire data frame and also specifies the class and levels of each element.
#### sapply() specifies the class and type of each individual element of teh data frame. For example, class(country) is "factor", typeof(country) is "interger"



## Step3: Part(a) Exploring individual variable-categorical; selecting Continent
#### Here I used a variable "a" to store the entire Continent's column.

```{r}
a <-select(gapminder, continent)
summary(a)
table(a)
class(a)

ggplot(gapminder) +
  stat_count(aes(continent, fill=continent))

```

#### The first output gives a brief summary of the variable. It comprises of 5 levels: Africa, Americas, Asia, Europe and Oceania.
#### table() generates the tabular form of the column
#### Combining ggplot with stats_count gives a distribution of continents indicating that Africa appeared the maximum number of times in the data set and Oceania appeared the least number of times.

## Step4: Part(b) Exploring individual variable-Quantitative: Selecting lifeExp
#### Here I used a variable "b" to store the entire lifeExp's column.

```{r}
b <- select(gapminder, lifeExp)
summary(b)
class(b)

ggplot(gapminder, aes(lifeExp)) +
  geom_histogram(aes(fill=continent))
```

#### The summary() function gives the distributions of lifeExp in the form of Mean, Median and Quartiles.
#### The histogram shows that lifeExp of Oceania is least. For the rest of the countries, lifeExp is spread out. 


## Step4: Exploring various Plots

#### Plots for two quantitative variables:

```{r}
gapminder %>% 
  filter(continent == "Europe" & country %in% c("Austria", "Denmark", "Portugal", "Sweden") ) %>% 
  ggplot(aes(year, lifeExp)) +
  geom_point(aes(color=country)) +
  geom_line(aes(color=country))
```
#### The above output shows that as lifeExp increases, gdpPercap also increases. Also, out of the above 4 countries in Europe, Portugal has shown the maximum change.


#### Plot for one Quantitative variable: lifeExp

```{r}
gapminder %>% 
  filter(continent == "Europe") %>% 
  ggplot(aes(factor(1), lifeExp)) +
  geom_violin(width = 0.6) +
  geom_jitter(aes(color=country), show.legend = FALSE, width = 0.2)
```

#### The plot shows that the average life expectancy in Europe is between 70-75 years.

#### Plot for one quantitative and one categorical variable

```{r}
gapminder %>% 
   filter(continent == "Europe" & country %in% c("Austria", "Denmark", "Portugal", "Sweden") ) %>% 
  ggplot(aes(country, gdpPercap)) +
  geom_boxplot(aes(color=country))+
  geom_jitter(width=0.3)
```

#### The output from the above plot shows that out of the 4 countries gdpPercap of Portugal is the least. One reason behind could be low life Expectancy.



## Step5: Other Explorations of ggplots()

#### Exploring Facetting

```{r}
gapminder %>% 
  filter(continent != "Europe") %>% 
  ggplot(aes(lifeExp, gdpPercap)) +
  scale_y_log10() +
  geom_point() +
  facet_wrap(~continent, nrow=4)

```
#### The above plots were obtained by using the facet_wrap() command which gave the output as 4 different plots each for a particular continent.The common conclusion from all the plots is that lifeExp and gdpPercap have a linear relationship that is as lifeExp increases, gdpPercap increases.


#### Exploring geom_smooth()

```{r}
gapminder %>% 
  filter(continent != "Europe") %>% 
  ggplot(aes(lifeExp, gdpPercap)) +
  scale_y_log10() +
  geom_point(alpha=0.35) +
  geom_smooth()
  
```

#### geom_smooth gives a smooth curve making data inference much easier. It can be concluded that if lifeExp is less, gdpPercap is low and vice-versa. 


#### Exploring dplyr
```{r}
gapminder %>% 
  mutate(ratio = (lifeExp/gdpPercap)) %>% 
  select(ratio, everything()) %>% 
  head

  
```
#### The above output was obtained using the dplyr function mutate(). It creates a new column by combining two or columns of the data frame. The first column of the output shows the ratio of lifeExp and gdpPercap.

```{r}
gapminder %>% 
  filter(continent == "Europe") %>% 
  summarise(avg_lifeExp = mean(lifeExp), avy_gdp = mean(gdpPercap))

```
#### The above output was obtained using the dplyr function summarise(). Stat function mean() gave the average lifeExp and gdpPercap of Europe as 71.9 dbl and 14469.5 dbl respectively.



#### Exploring Tables

#### Loading knitr and kable into R markdown

```{r}
library(knitr)
library(kableExtra)

```


#### Creating table using the first 6 rows and columns and exploring few some styling options

```{r}
g <- gapminder[1:6,1:6]
  kable(g) %>% 
  kable_styling(bootstrap_options = c("striped")) %>% 
  row_spec(1:6, background = "yellow")


```

## Step6: Commit and Push to github Repo
